First Order Alternation

### 2019-02-03 quantifiers/body.tex
ok

Many results in formal language theory rely on the assumption that
languages are defined over finite alphabets. In practice, this
assumption is problematic when attempting to use automata as models of
real-time systems or even simple programs, whose input and observable
output requires taking into account data values, ranging over very
large domains, better viewed as infinite mathematical abstractions.

Alternating automata are a generalization of nondeterministic automata
with universal transitions, that create several copies of the
automaton, which synchronize on the same input word. Alternating
automata are appealing for verification because they allow encoding of
problems such as temporal logic model checking in linear time, as
opposed to the exponential time required by nondeterministic automata
\cite{Vardi95}. A finite-alphabet alternating automaton is typically
described by a set of transition rules $q \arrow{a}{} \phi$, where $q$
is a state, $a$ is an input symbol and $\phi$ is a positive boolean
combinations of states, viewed as propositional variables.

Here we introduce a generalized alternating automata model in which
states are predicate symbols $q(y_1,\ldots,y_k)$, the input has
associated data variables $x_1,\ldots,x_n$, ranging over an infinite
domain and transitions are of the form $q(y_1,\ldots,y_k)
\arrow{a(x_1,\ldots,x_n)}{} \phi$, where $\phi$ is any formula in the
first-order theory of the data domain, in which each state predicate
occurs under an even number of negations. In this model, the arguments
of a predicate atom $q(y_1,\ldots,y_k)$ track the values of the
\emph{internal variables} associated with the state. Together with the
input values $x_1,\ldots,x_n$, these values are used to compute the
successor states and are invisible in the input sequence.

Previous attempts to generalize classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory (register) automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$.

The solution we adopt here is a tight coupling of internal variables
to control states, using uninterpreted predicate symbols. As we show,
this allows for linear-time complementation just as in the case of
boolean alternating automata. Complementation is, moreover, possible
when the transition formulae contain first-order quantifiers,
generating infinitely-branching execution trees. The price to be paid
for this expressivity is that emptiness of first-order alternating
automata is undecidable, even for the simplest data theory of equality
\cite{Farzan15}.

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~procedure, originally developed for checking safety of
nondeterministic integer programs \cite{McMillan06}. However, checking
emptiness of first-order alternating automata by lazy annotation with
interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, or even impossible, to
  decide if a given symbolic trace is spurious. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the spuriousness of a symbolic path
  are not \emph{local}, as they may refer to input values encountered
  in the past. However, the future executions are oblivious to
  \emph{when} these values have been seen in the past and depend only
  on the data constraints between the values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

As applications of first order alternating automata, we identified
several undecidable problems for which no semi-algorithmic methods
exist: inclusion between recognizable timed languages
\cite{AlurDill94}, languages recognized by finite-memory automata
\cite{KaminskiFrancez94} and emptiness of predicate automata, a
subclass of first-order alternating automata used to check safety and
liveness properties of parameterized concurrent programs
\cite{Farzan15,Farzan16}.

For reasons of space, all proofs of technical results in this paper
are given in \cite{IosifXu18ArXiv}.

\paragraph{\bf Related Work} 
The first order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered states to be propositional variables, as in the classical
textbook alternating automata model, and all variables of the
automaton to be observable in the input. The model in this paper
overcomes this latter restriction by allowing for internal variables,
whose variables are not visible in the language.

>>>>

Many results in automata theory rely on the finite alphabet
hypothesis, which guarantees, in some cases, the existence of
determinization, complementation and inclusion checking
methods. However, this hypothesis prevents the use of automata as
models of real-time systems or even simple programs, whose input and
output are data values ranging over very large domains, typically
viewed as infinite mathematical abstractions.

Traditional attempts to generalize classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$.

Even when assuming the alphabet to be finite, complementation of a
finite-state automaton faces an inherent exponential blowup due to
nondeterminism. However, if we allow universal nondeterminsim, in
addition to the classical existential nondeterminism, complementation
is possible is linear time. The automata having both existential and
universal nondeterminism are called \emph{alternating automata}
\cite{ChandraKozenStockmeyer81}. A finite-alphabet alternating
automaton is described by a set of transition rules $q \arrow{a}{}
\phi$, where $q$ is a state, $a$ is an input symbol and $\phi$ is a
boolean formula, whose propositional variables denote successor
states.

\paragraph{\em Our Contribution}
We extend the alternating automata model to infinite data alphabets,
by defining a model of computation in which all boolean operations,
including complementation, can be done in linear time.  The control
states are given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the
input consists of an event $a$ from a finite alphabet and a tuple of
data variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
transitions are of the form $q(y_1,\ldots,y_k)
\arrow{a(x_1,\ldots,x_n)}{} \phi(x_1, \ldots, x_n, y_1, \ldots, y_k)$,
where $\phi$ is a formula in the first-order theory of the data
domain. In this model, the arguments of a predicate atom
$q(y_1,\ldots,y_k)$ represent the values of the \emph{internal
  variables} associated with the state. Together with the input values
$x_1,\ldots,x_n$, these values define the next configurations, but
remain invisible in the input sequence.

The tight coupling of internal values and control states, by means of
uninterpreted predicate symbols, allows for linear-time
complementation just as in the case of classical propositional
alternating automata. Complementation is, moreover, possible when the
transition formulae contain first-order quantifiers, generating
infinitely-branching execution trees. The price to be paid for this
expressivity is that emptiness of first-order alternating automata is
undecidable, even for the simplest data theory of equality
\cite{Farzan15}.

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word $w \in \lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches.

Checking emptiness of first-order alternating automata by lazy
annotation with interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, to
  decide if a path is infeasible. This is mainly because adding
  uninterpreted predicate symbols to decidable first-order theories,
  such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

\paragraph{\em Applications}
We use first order alternating automata to develop practical
semi-algorithms for a number of known undecidable problems, such as:
inclusion of regular timed languages \cite{AlurDill94}, inclusion of
quasi-regular languages recognized by finite-memory automata
\cite{KaminskiFrancez94} and emptiness of predicate automata, a
subclass of first-order alternating automata used to verify
parameterized concurrent programs \cite{Farzan15,Farzan16}.

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event occurs after at least $2$ time units,
whereas $B$ recognizes the sequences in which there exists an event
$a$ such that no event occurs after it in exactly $1$ time
unit. Formally, $\lang{A} = \set{a, t_1, \ldots a, t_n \mid t_1,
  \ldots, t_n \in \real,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1,
  t_1 \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ t_1, \ldots, t_n \in
  \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
  \cite[Theorem 1]{AlurMadhusudan04} for a proof.} that $\clang{B}$ is
not recognized by a timed automaton, thus the inclusion cannot be
proved in the framework of timed regular languages, as $\lang{A} \cap
\clang{B} = \emptyset$.

We encode timed automata as first order alternating automata and uses
closure of the latter class under intersection and complement to
compute an alternating automaton whose emptiness proves the inclusion
$\lang{A} \subseteq \lang{B}$ between timed regular languages. The
idea is to represent each control state of $A$ and $B$ by a binary
predicate symbol with the same name. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}. The first argument of
$s_1(x,z)$ is used to track the value of the $x$ clock of $A$, whereas
$z$ is the current time value. The time stamp of the input event is
the input variable $t$, which is bigger or equal than the current time
$z$, since the lapse of time between two consequtive events is always
positive. The current value of the clock is $z-x$, thus the guard $x
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule. Note that the time increment
equals the increment of the clocks that are not reset.

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{tabular}{ccc}
\begin{minipage}{3cm}
\input ta.pdf_t
\end{minipage} 
&&
\begin{minipage}{5cm}
\small{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]}
\end{minipage} \\
&& 
\begin{minipage}{5cm}
{\small\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{tabular}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom
right). Finally, we check the emptiness of the alternating automaton
consisting of the rules of $A$ and $\overline{B}$, started in the
initial configuration $s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation
semi-algorithm finds that this automaton is empty in \comment[ri]{...}
seconds.

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered states to be propositional variables, as in the classical
textbook alternating automata model, and all variables of the
automaton to be observable in the input. The model in this paper
overcomes this latter restriction by allowing for internal variables,
whose variables are not visible in the language.
